Variational Monte Carlo on Bosonic Systems

### 2018-04-06 project_1/article/article.tex
Explain look of steepest descent figure
Spell check results on non-interaction

            \item A ratio, $q$, between the probability density of the new
                wavefunction and the previous probability density is computed.
                We accept the step if $q \geq 1$, else we accept the step with a
                probability $p$. The latter is done by drawing a random uniform
                number, $x \in [0, 1]$, and checking if $q > x$.  This
                acceptance probability decides if the particle move is rejected
                or accepted.

@@

        The algorithm described above can be applied in an "exhaustive" search
        of the parameter space in order to find the optimal parameters.  Whether
        a proposed move is accepted or not is determined by a \emph{transition
        probability}. The strength of the algorithm lies in that the transition
        algorithm need not be known. For example, the simplest case is to accept
        the new state, i.e., the new position for the random walker, if the
        ratio
        \begin{align}
            q(\vf{r}_{i + 1}, \vf{r}_i)
            &=
            \frac{\left|\Psi_T(\vf{r}_{i + 1})\right|^2}
            {\left|\Psi_T(\vf{r}_{i})\right|^2},
        \end{align}
        where $\vf{r}_{i + 1}$ are all the positions at step $i + 1$, is greater
        than a uniform probability $p \in [0, 1]$. This transition criteria
        results in what we call the brute force Metropolis-Hastings algorithm.
        By counting the number of accepted steps and dividing by the total
        number of Monte Carlo cycles we get the \emph{acceptance probability}.
        The acceptance probability helps in choosing a decent length for the
        proposed new step. If the acceptance probability is very low we can
        lower the step length to get more accepted steps and vice versa.

@@

                This is a so-called "hacky" solution, but it does the trick. It
                is also the reason why the result figure in
                \autoref{fig:gradient_descent_interacting}.

@@

    \subsection{Non-Interacting Systems}
        Initially we are interested in seeing how our implementation performs
        and to what extent it functions properly.  the logical thing to do would
        be to look at a non-interacting system with a simple Gaussian
        wavefunction and harmonic oscillator potential. Such a system has a
        simple closed-form analytic solution\footnote{A fact that every
        undergraduate physics student should be well aware of after finishing
        their first course in quantum physics}, which is further simplified by
        the "god-given" analytic units
        (\autoref{eq:closed_form_natural_units_local_energy}). This expression
        is minimised for $\alpha = \frac{1}{2}\omega$. We also set $\omega = 1$
        and get an expression for the expected minimum local energy,
        \begin{equation}
            E_L(\vb{r}) = \frac{dN}{2},
        \end{equation}
        which will be used as the benchmark we are ultimately aiming for.

        We will start without importance sampling, that is by employing the
        simpler and more naïve "brute force" Monte Carlo sampling. Moreover, we
        will compute the value for the local energy analytically as well as with
        a numerical approach.  In the numerical approach we employ the well-know
        approximation for the second derivative,
        \begin{equation}
                f''(x) \approx \frac{f(x-h) - 2f(x) + f(x+h)}{h^2}.
        \end{equation}

        \begin{table}
            \centering
            \caption{One particle in one dimension for the analytic expression
            with $2^{21}$ Monte Carlo cycles and step length of $0.5$.}
            \begin{ruledtabular}
                \begin{tabular}{cccccc}
                    $\alpha$ & $\langle  E_L\rangle$ & $\sigma$ & $\sigma_b$
                    & $A$ & $t_{\text{CPU}} [\si{\second}]$ \\
                    \hline
                    0.30&0.56244&0.00026&0.00129&0.892&0.43\\
                    0.34&0.53652&0.00019&0.00094&0.885&0.38\\
                    0.38&0.51863&0.00013&0.00063&0.878&0.37\\
                    0.42&0.50805&0.00009&0.00038&0.872&0.38\\
                    0.46&0.50201&0.00004&0.00017&0.866&0.37\\
                    0.50&0.50000&0.00000&0.00000&0.860&0.37\\
                    0.54&0.50183&0.00004&0.00015&0.855&0.38\\
                    0.58&0.50566&0.00007&0.00029&0.850&0.38\\
                    0.62&0.51175&0.00011&0.00041&0.845&0.37\\
                    0.66&0.51906&0.00014&0.00052&0.840&0.37\\
                    0.70&0.52883&0.00017&0.00062&0.836&0.37\\
                \end{tabular}
            \end{ruledtabular}
            \label{tab:1D1N_analytic}
        \end{table}

@@

        \autoref{tab:1D1N_analytic} shows the expected local energy
        $\expval{E_L}$, standard deviations $\sigma$, $\sigma_b$, the share of
        accepted Monte Carlo reorientations $A$ and cpu time for the analytic
        computation at different values for the variational parameter $\alpha$.
        \autoref{tab:1D1N_numeric} shows the same numbers but for the numeric
        scheme. The first column of standard deviations $\sigma$ are the "naïve"
        standard deviations, while $\sigma_b$ marks the standard deviations
        computed with the blocking method. One can clearly see, in both the
        analytic and numeric case, that the former standard deviations are lower
        than the latter, meaning that autocorrelated data can lead to a lower
        perceived uncertainty than what is reality. The blocking method takes
        this fact into account and for this reason, all other standard
        deviations will be computed with the blocking method from here on.

        The first initial observation of note is that we have found an optimum
        for the system in terms of a minimum energy at $\alpha=\frac{1}{2}$.
        Moreover, notice how the standard deviation disappears completely.  This
        should already be apparent from
        \autoref{eq:closed_form_natural_units_local_energy}, as the term
        involving particle positions disappears for the optimal $\alpha$.

        \begin{table}
            \caption{Comparison of energy, standard deviation and the CPU time
            for $\alpha = 0.5$ (the correct minimum) for the analytic and
            numeric schemes in three dimensions and $N$ particles. The step
            length is $0.5$.}
            \centering
            \begin{ruledtabular}
                \begin{tabular}{r|rrr|rrr}
                    & \multicolumn{3}{c|}{Analytic}
                    & \multicolumn{3}{c}{Numeric} \\
                    \hline
                    $N$
                    & $\langle E_L\rangle$ & $\sigma_b$
                    & $t_{\text{CPU}} [\si{\second}]$
                    & $\langle E_L\rangle$ & $\sigma_b$
                    & $t_{\text{CPU}} [\si{\second}]$ \\
                    \hline
                    1 & 1.5 & 0.0 & 0.50
                    & 1.49997 & 0.00008 & 0.73 \\
                    10 & 15.0 & 0.0 & 0.73
                    & 14.98278 & 0.00867 & 6.40 \\
                    100 & 150.0 & 0.0 & 4.13
                    & 149.81453 & 0.81501 & 387.60 \\
                    500 & 750.0 & 0.0 & 19.16
                    & 773.85435 & 20.69789 & 9595.02
                \end{tabular}
            \end{ruledtabular}
            \label{tab:initial_optimum_comparison}
        \end{table}

        Systems of optimum variational parameter warrants further investigation,
        as summed up in \autoref{tab:initial_optimum_comparison}. The analytical
        results are quite predictable and monotonous; the expected local energy
        $\expval{E_L}$ is perfectly proportional by a factor $\frac{3}{2}$ to
        the number of particles and the standard deviation $\sigma_b$ is zero
        for systems of any number of particles. The numerical results tells a
        more tortuous tale. As the number of particles increase, the uncertainty
        also increases rapidly. For the largest system, with $N = 500$
        particles, the energy misses its target critically - the true energy is
        more than one standard deviation away. Lastly, the cpu time is notably
        higher for the numeric scheme.

@@

\clearpage
\onecolumngrid
\section{Brute Force Metropolis-Hastings}
    \begin{figure}[b]
        \makebox[\textwidth][c]{
            \includegraphics[width=1.2\textwidth]{../data/figures/problem_b.pdf}
        }
        \caption{Comparison of analytic and numeric results with the exact
        expression for the energy as a function of $\alpha$. The plots show the
        standard deviation from the blocking method as error ticks. In the
        figure $N$ is the number of particles and $d$ the number of dimensions.}
        \label{fig:initial_problem_b}
        \vspace{2cm}
    \end{figure}

>>>>

            \item A ratio, $q$, between the probability density of the new
                wavefunction and the previous probability density is computed.
                We accept the step if $q \geq 1$, else we accept the step with a
                probability $p$. The latter is done by drawing a random uniform
                number, $x \in [0, 1]$, and checking if $q > x$.  This
                transition probability decides if the particle move is rejected
                or accepted.

@@

        The algorithm described above can be applied in an "exhaustive" search
        of the parameter space in order to find the optimal parameters.  Whether
        a proposed move is accepted or not is determined by a \emph{transition
        probability}. The strength of the algorithm lies in that the transition
        algorithm need not be known. For example, the simplest case is to accept
        the new state, i.e., the new position for the random walker, if the
        ratio
        \begin{align}
            q(\vf{r}_{i + 1}, \vf{r}_i)
            &=
            \frac{\left|\Psi_T(\vf{r}_{i + 1})\right|^2}
            {\left|\Psi_T(\vf{r}_{i})\right|^2},
        \end{align}
        where $\vf{r}_{i + 1}$ are all the positions at step $i + 1$, is greater
        than a uniform probability $p \in [0, 1]$. This transition criteria
        results in what we call the brute force Metropolis-Hastings algorithm.
        By counting the number of accepted steps and dividing by the total
        number of Monte Carlo cycles we get the \emph{acceptance probability},
        denoted $A$.  The acceptance probability helps in choosing a decent
        length for the proposed new step. If the acceptance probability is very
        low we can lower the step length to get more accepted steps and vice
        versa.

@@

                This is a so-called "hacky" solution, but it does the trick. It
                is also the reason why the resultant figure in
                \autoref{fig:gradient_descent_interacting} looks the way it
                does for $\alpha_0 = 0.2$ and $\alpha_0 = 0.3$.

@@

    \subsection{Non-Interacting Systems}
        Initially we are interested in seeing how our implementation performs
        and to what extent it functions properly. The logical thing to do would
        be to look at a non-interacting system with a simple Gaussian
        wavefunction and harmonic oscillator potential. Such a system has a
        simple closed-form analytic solution\footnote{A fact that every
        undergraduate physics student should be well aware of after finishing
        their first course in quantum physics}, which is further simplified by
        the "god-given" analytic units
        (\autoref{eq:closed_form_natural_units_local_energy}). This expression
        is minimised for $\alpha = \omega/2$. We also set $\omega = 1$
        and get an expression for the expected minimum local energy,
        \begin{equation}
            E_L(\vb{r}) = \frac{dN}{2},
        \end{equation}
        which will be used as the benchmark we are ultimately aiming for.

        We will start without importance sampling, that is by employing brute
        force Monte Carlo sampling. Moreover, we will compute the value for the
        local energy analytically as well as with a numerical approach.  In the
        numerical approach we employ the second order central difference
        approximation
        \begin{equation}
            \dod[]{f(x)}{x} \approx \frac{f(x-h) - 2f(x) + f(x+h)}{h^2}.
        \end{equation}

        \begin{table}
            \centering
            \caption{One particle in one dimension for the analytic expression
            with $2^{21}$ Monte Carlo cycles and a step length of $0.5$.}
            \begin{ruledtabular}
                \begin{tabular}{cccccc}
                    $\alpha$ & $\langle  E_L\rangle$ & $\sigma$ & $\sigma_b$
                    & $A$ & $t_{\text{CPU}} [\si{\second}]$ \\
                    \hline
                    0.30&0.56244&0.00026&0.00129&0.892&0.43\\
                    0.34&0.53652&0.00019&0.00094&0.885&0.38\\
                    0.38&0.51863&0.00013&0.00063&0.878&0.37\\
                    0.42&0.50805&0.00009&0.00038&0.872&0.38\\
                    0.46&0.50201&0.00004&0.00017&0.866&0.37\\
                    0.50&0.50000&0.00000&0.00000&0.860&0.37\\
                    0.54&0.50183&0.00004&0.00015&0.855&0.38\\
                    0.58&0.50566&0.00007&0.00029&0.850&0.38\\
                    0.62&0.51175&0.00011&0.00041&0.845&0.37\\
                    0.66&0.51906&0.00014&0.00052&0.840&0.37\\
                    0.70&0.52883&0.00017&0.00062&0.836&0.37\\
                \end{tabular}
            \end{ruledtabular}
            \label{tab:1D1N_analytic}
        \end{table}

@@

        \autoref{tab:1D1N_analytic} shows the expected local energy
        $\expval{E_L}$, the uncorrelated standard deviation, $\sigma$, the
        standard deviation found from blocking, $\sigma_b$, the accpetance
        probability, $A$, and the CPU time for the analytic computation at
        different values for the variational parameter $\alpha$.
        \autoref{tab:1D1N_numeric} shows the same numbers but for the numeric
        scheme.  One can clearly see, in both the analytic and numeric case,
        that the uncorrelated standard deviations are lower than the standard
        deviations found from the blocking method. This means that
        autocorrelated data can lead to a lower perceived uncertainty than what
        is the reality. The blocking method takes this fact into account and for
        this reason, all other standard deviations will be computed with the
        blocking method from here on.

        The first initial observation of note is that we have found an optimum
        for the system in terms of a minimum energy at $\alpha = 1/2$.
        Moreover, notice how the standard deviation disappears completely.  This
        should already be apparent from
        \autoref{eq:closed_form_natural_units_local_energy}, as the term
        involving particle positions disappears for the optimal $\alpha$ thus
        rendering the entire Monte Carlo process of moving particles void.

        \begin{table}
            \caption{Comparison of energy, standard deviation and the CPU time
            for $\alpha = 0.5$ for the analytic and numeric schemes in three
            dimensions and $N$ particles. The step length is $0.5$.}
            \centering
            \begin{ruledtabular}
                \begin{tabular}{r|rrr|rrr}
                    & \multicolumn{3}{c|}{Analytic}
                    & \multicolumn{3}{c}{Numeric} \\
                    \hline
                    $N$
                    & $\langle E_L\rangle$ & $\sigma_b$
                    & $t_{\text{CPU}} [\si{\second}]$
                    & $\langle E_L\rangle$ & $\sigma_b$
                    & $t_{\text{CPU}} [\si{\second}]$ \\
                    \hline
                    1 & 1.5 & 0.0 & 0.50
                    & 1.49997 & 0.00008 & 0.73 \\
                    10 & 15.0 & 0.0 & 0.73
                    & 14.98278 & 0.00867 & 6.40 \\
                    100 & 150.0 & 0.0 & 4.13
                    & 149.81453 & 0.81501 & 387.60 \\
                    500 & 750.0 & 0.0 & 19.16
                    & 773.85435 & 20.69789 & 9595.02
                \end{tabular}
            \end{ruledtabular}
            \label{tab:initial_optimum_comparison}
        \end{table}

        Systems of optimum variational parameter warrants further investigation,
        as summed up in \autoref{tab:initial_optimum_comparison}. The analytical
        results are quite predictable and monotonous; the expected local energy
        $\expval{E_L}$ is perfectly proportional by a factor $3/2$ to
        the number of particles and the standard deviation $\sigma_b$ is zero
        for systems of any number of particles. The numerical results tells a
        more tortuous tale. As the number of particles increase, the uncertainty
        also increases rapidly. For the largest system, with $N = 500$
        particles, the energy misses its target critically - the true energy is
        more than one standard deviation away. Lastly, the CPU time is notably
        higher for the numeric scheme.

@@

\clearpage
\onecolumngrid
\section{Brute Force Metropolis-Hastings}
    In \autoref{fig:initial_problem_b} we see how the expectation value of the
    energy for the analytic and numeric expression of the local energy using
    Monte Carlo sampling matches the exact energy from \autoref{eq:exact_energy}
    for $\alpha \in [0.3, 0.7]$.
    \begin{figure}[b]
        \makebox[\textwidth][c]{
            \includegraphics[width=1.2\textwidth]{../data/figures/problem_b.pdf}
        }
        \caption{Comparison of analytic and numeric results with the exact
        expression for the energy as a function of $\alpha$. The plots show the
        standard deviation from the blocking method as error ticks. In the
        figure $N$ is the number of particles and $d$ the number of dimensions.}
        \label{fig:initial_problem_b}
        \vspace{2cm}
    \end{figure}
